$\pi$ with leftovers: \\ a mechanisation in Agda

### 2021-01-27 paper/paper.tex
Fix spacing

\begin{nidefinition}[\textsc{StructCong}] \label{def:structcong}
We define the base cases of a structural congruence relation $\eqeq$ as follows:

\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : \Set}
  \; \textsc{StructCong}

@@

  \inferrule
  { }
  {\constr{scope-comm} : \new \new P \eqeq \new \new \func{exchange}_{\constr{0}} \; P}
\end{mathpar}

The first three rules ($\constr{comp}-$*) state associativity, symmetry, and $\PO$ as being the neutral element of parallel composition, respectively.
The last three ($\constr{scope}-$*) state garbage collection, scope extrusion and commutativity of restrictions, respectively.
In $\constr{scope-ext}$ the side condition $\Unused_i \; Q$ makes sure that $i$ is unused in $Q$ (see \autoref{def:unused}).
The function $\func{lower}_i \; Q \; uQ$ traverses $Q$ decrementing every index greater than $i$.
In $\constr{scope-comm}$ the function $\func{exchange}_i \; P$ traverses $P$ (of type $\allowbreak \Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all the above, $i$ is incremented every time we go under a binder.
\end{nidefinition}

@@

\begin{nidefinition}[\textsc{Algebra}]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning) and \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition).
  In addition, we ask that the leftovers can be \emph{computed} so that we can automatically update the usage context every time input and output occurs --- this is purely for usability.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  Below we define such an algebra as a record $\Algebra_C$ on a carrier $C$.
  (We use $\forall$ for universal quantification.
  The dependent product $\type{\exists}$ uses the value of its first argument in the type of its second.
  The type $\type{DEC} \; P$ is a witness of either $P$ or $P \to \type{\bot}$, where $\type{\bot}$ is the empty type with no constructors.)

  \begin{equation*}
  \begin{aligned}
    &\op{\_}{\_}{\_}         &:{} &                 &        & C \to C \to C \to \Set \\
    &\func{\cdot-unique}     &:{} &\forall x x' y z  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\func{\cdot-unique^l}   &:{} &\forall x y y' z  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\func{\cdot-assoc}      &:{} &\forall x y z u v & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w  \; (\op{x}{u}{w} \times \op{w}{v}{z}) \\
    &\func{\cdot-comm}       &:{} &\forall x y z     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\func{\cdot-compute^r}  &:{} &\forall x y       & \to \; & \type{DEC} \; (\type{\exists} z  \; (\op{x}{y}{z})) \\
    &\zero                   &:{} &                 &        & C \\
    &\func{\cdot-id^l}       &:{} &\forall x         & \to \; & \op{x}{\zero}{x} \\
    &\func{\cdot-min^l}      &:{} &\forall y z        & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\one                    &:{} &                 &        & C \\
  \end{aligned}
  \end{equation*}
\end{nidefinition}

@@

\begin{nidefinition}[\textsc{Types}: typing processes] \label{def:types}
  The \textsc{Types} typing relation for the resource-aware \picalc{} processes is presented below.
  For convenience, we reuse the constructor names introduced for the syntax in \autoref{syntax}.

  \begin{mathpar}
    \datatype{
      \gamma : \PreCtx_n \\
      P : \Process_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma \; \Delta : \Ctx_{idxs}}}
    {\types{\gamma}{\Gamma}{P}{\Delta} : \Set}
    \; \textsc{Types}

>>>>

\begin{nidefinition}[\textsc{StructCong}] \label{def:structcong}
We define the base cases of a structural congruence relation $\eqeq$ as follows:
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : \Set}
  \; \textsc{StructCong}

@@

  \inferrule
  { }
  {\constr{scope-comm} : \new \new P \eqeq \new \new \func{exchange}_{\constr{0}} \; P}
\end{mathpar}
The first three rules ($\constr{comp}-$*) state associativity, symmetry, and $\PO$ as being the neutral element of parallel composition, respectively.
The last three ($\constr{scope}-$*) state garbage collection, scope extrusion and commutativity of restrictions, respectively.
In $\constr{scope-ext}$ the side condition $\Unused_i \; Q$ makes sure that $i$ is unused in $Q$ (see \autoref{def:unused}).
The function $\func{lower}_i \; Q \; uQ$ traverses $Q$ decrementing every index greater than $i$.
In $\constr{scope-comm}$ the function $\func{exchange}_i \; P$ traverses $P$ (of type $\allowbreak \Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all the above, $i$ is incremented every time we go under a binder.
\end{nidefinition}

@@

\begin{nidefinition}[\textsc{Algebra}]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning) and \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition).
  In addition, we ask that the leftovers can be \emph{computed} so that we can automatically update the usage context every time input and output occurs --- this is purely for usability.
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  Below we define such an algebra as a record $\Algebra_C$ on a carrier $C$.
  (We use $\forall$ for universal quantification.
  The dependent product $\type{\exists}$ uses the value of its first argument in the type of its second.
  The type $\type{DEC} \; P$ is a witness of either $P$ or $P \to \type{\bot}$, where $\type{\bot}$ is the empty type with no constructors.)
  \begin{equation*}
  \begin{aligned}
    &\op{\_}{\_}{\_}         &:{} &                 &        & C \to C \to C \to \Set \\
    &\func{\cdot-unique}     &:{} &\forall x x' y z  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\func{\cdot-unique^l}   &:{} &\forall x y y' z  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\func{\cdot-assoc}      &:{} &\forall x y z u v & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w  \; (\op{x}{u}{w} \times \op{w}{v}{z}) \\
    &\func{\cdot-comm}       &:{} &\forall x y z     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\func{\cdot-compute^r}  &:{} &\forall x y       & \to \; & \type{DEC} \; (\type{\exists} z  \; (\op{x}{y}{z})) \\
    &\zero                   &:{} &                 &        & C \\
    &\func{\cdot-id^l}       &:{} &\forall x         & \to \; & \op{x}{\zero}{x} \\
    &\func{\cdot-min^l}      &:{} &\forall y z        & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\one                    &:{} &                 &        & C \\
  \end{aligned}
  \end{equation*}
\end{nidefinition}

@@

\begin{nidefinition}[\textsc{Types}: typing processes] \label{def:types}
  The \textsc{Types} typing relation for the resource-aware \picalc{} processes is presented below.
  For convenience, we reuse the constructor names introduced for the syntax in \autoref{syntax}.
  \begin{mathpar}
    \datatype{
      \gamma : \PreCtx_n \\
      P : \Process_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma \; \Delta : \Ctx_{idxs}}}
    {\types{\gamma}{\Gamma}{P}{\Delta} : \Set}
    \; \textsc{Types}
